Introductory statistics with `intRo`

### 2016-06-06 jcgs-paper-2016/paper-original.Rmd
Rename to original

Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. This in turn led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of use and an aim to foster student interest in coding. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text. The paper is structured as follows: section \ref{what-is-intro} introduces the application, its features and its usability, and provides motivations for the design decisions underlying it. Section \ref{understanding-intro} provides technical details on the design decisions of `intRo`, in particular, its modular and extensible structure. Section \ref{deploying-intro-instances} describes the process of deploying a classroom instance of `intRo` tailored to the individual instructor. Finally, section \ref{conclusions-and-future-work} discusses some future possibilities and limitations of the software.

@@

In the development of `intRo`, we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find what they need more easily. We also made the functionality available the minimal necessary for an introductory statistics course. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in `intRo`. In this instance, a user clicks on the Graphical tab to create a mosaic plot. The user sees the plot, and elects to click the save button to store the plot and its corresponding code to the final compendium.

@@

We've also created a documentation website that is consistent with the interface of the application. The documentation is available at http://gammarama.github.io/intRo and hyperlinked within the application itself. The documentation describes the specific functionality provided by the default `intRo` modules.

Statistics courses with a lab component can most benefit from the use of `intRo`. As students learn the aforementioned concepts, they can practice using the tools to perform the statistical tests and summaries. They simply upload a dataset and click the module appropriate to the section of the class. Once they choose the options desired by the instructor (variables, confidence levels, etc.), they can view the results and ultimately print the output for turning in. Early labs can walk the user through the limited number of button presses and clicks necessary, while later labs can more vaguely describe the intended output in hopes that the user can easily navigate the `intRo` interface to obtain the results.

`intRo` has an ulterior motive as well: to get students excited about programming. By navigating about the user interface of `intRo`, students are actually creating a fully-executable R script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of `intRo` is meant to generate excitement about programming in R and empower students to feel that they can generate code as well. `intRo` uses rmarkdown's render function in order to print the results, by dymanically executing the user's R script. By default, the output will include the R code, but if the user elects to hide the source code by clicking the code toggle button at the top, the code will not appear in the printed results. This reproducibility framework has a number of key advantages:

On the front end, user interaction with `intRo` is split into bitesize chunks that we call modules. Modules are self-contained pieces of functionality which implement common statistical procedures that are often used in introductory statistics classes. These modules form the core of `intRo` and are discussed at length in the next section.

To fully understand `intRo`, it is important to expand upon two underlying concepts. The first is the idea of modularity in the context of a Shiny application. The second is the idea of reactive programming in the context of a web application.

@@

`intRo`'s user interface and functionality is dynamically generated depending on the set of modules enabled. This process is detailed in figure \ref{fig:app_creation_modules}. The key driver to populating \texttt{server.R} and \texttt{ui.R} is the navigation file \texttt{modules.txt}, which defines the placement of each module. The interface is then created with the following statement.

@@

@@

This script is most immediately similar to standard R code. In this case, a function `cont.table` is created which, depending on the values of the parameters, ultimately returns a contingency table. One important difference from a typical R script is that each call in the script is wrapped in a function called `interpolate` [@interpolate]. `interpolate` both executes the given R code on the server, and also writes the code executed to the script window at the bottom of `intRo`.

@@

@@

# Deploying `intRo` Instances

Although users can access intRo from \url{http://www.intro-stats.com}, course instructors may wish to download, customize, and deploy their own instance, perhaps with new modules or modified theming or functionality. `intRo` can be downloaded, ran, and deployed on ShinyApps.io through the use of the R package `intRo`. Currently, the package is only available on GitHub, and can be installed using the devtools package as follows:

If the intent is to use a specific instance of `intRo` where many users will access it at the same time, such as in an introductory statistics class, it may be preferable to deploy a custom instance of `intRo` to a publicly accessible URL. The package provides a function `deploy_intRo` which is a wrapper for the `deployApp` function contained in the shinyapps package. Once the shinyapps package is installed and configured, `deploy_intRo` will upload `intRo` as an application on the user's ShinyApps.io account. The function takes the same arguments as `run_intRo`, so it can be deployed with a custom selection of modules, and a customized theme. It also takes an additional argument `google_analytics`, which allows the specification of a Google Analytics tracking ID. It also takes `...` as additional arguments to be passed into the `deployApp` routine. For example, if we wished to deploy the instance of `intRo` we ran previously, we would call it like so:

@@

# Supplementary material

All code and documents related to this manuscript are available at \url{https://github.com/gammarama/intRo}.

>>>>

Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. This in turn led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of use and an aim to foster student interest in coding. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text. The paper is structured as follows: section \ref{what-is-intro} introduces the application, its features and its usability. Section \ref{using-intro} describes the process of creating a classroom instance of `intRo` tailored to the individual instructor. Section \ref{understanding-intro} provides technical details on the implementation of `intRo`. Finally, section \ref{conclusions-and-future-work} discusses some future possibilities and limitations of the software.

@@

In the development of `intRo`, we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find what they need more easily. We also made the functionality available the minimal necessary for an introductory statistics course. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in `intRo`. In this instance, a user clicks on the Graphical tab to create a mosaic plot. The user sees the plot, and elects to click the save button to store the plot (and its corresponding code) to the final compendium.

@@

We've also created a documentation website that is consistent with the interface of the application. The documentation is available at http://gammarama.github.io/intRo and hyperlinked within the application itself.

`intRo` has an ulterior motive as well: to get students excited about programming. By navigating about the user interface of `intRo`, students are actually creating a fully-executable R script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of `intRo` is meant to generate excitement about programming in R and empower students to feel that they can generate code as well. `intRo` uses rmarkdown's render function in order to print the results, by dymanically executing the user's R script. By default, the output will include the R code, but if the user elects to hide the source code by clicking the code toggle button at the top, the code will not appear in the printed results.

On the front end, user interaction with `intRo` is split into bitesize chunks that we call modules. Modules are self-contained pieces of functionality which implement common statistical procedures that are often used in introductory statistics classes.

Within the data category there are two modules, sources and transformation. The data sources module allows the user to select or upload a dataset. Several datasets are pre-included, which can be selected via a drop-down menu. The user can check a box to upload a dataset, and then select the CSV file from her computer. Upon selection or successful upload of a dataset, a table view of the data will be visible on the right. From this module the user can select a subset of the data either by value or random selection. The dataset selected in the data module will be made available to all other modules of `intRo`. The data transformation module allows transformation of numeric variables to categorical and categorical variable to numeric. Additionally, numeric variable can be transformed using a power transformation. The transformed variables can be saved to the dataset for use elsewhere in `intRo`.

The graphical summaries module provides several common graphical displays of variables in the dataset. First, the user selects the plot desired, and either a single variable or two variables from the dataset, depending on plot type. The variable available to choose will only be those compatible with plot type selected. The plot visible on the right will automatically update when these values are selected. The numeric summaries module is another method of computing summaries of the data. The user is presented with a list of each variable from their dataset. The user can select one or more different variables, and will be presented with a table of some summary statistics. The summary statistics displayed vary depending on the type of the variables. For instance, for a numeric variable, a five number summary will be displayed. Character or factor variables will display counts of each unique value of that particular variable. Additionally, the user has the ability to display summaries by a grouping variable.

Within the set of statistics modules, there are three simple statistics tasks that an introductory statistics student will learn in her first semester. The contingency table module allows the user to compute a contingency table for two categorical variables. The results can either be displayed as counts, row percentages, column percentages, or total percentages. The simple linear regression module allows the user to compute a simple linear regression of a dependent variable (y) on an independent variable (x). The module will automatically filter any variables that are non-numeric so that the user may only select variables that are suitable for a simple linear regression. Results of the regression will be displayed on the right. A table of parameter estimates, a scatterplot of the data along with a line of best fit, The R-Squared value, and residual diagnostic plots will all be displayed. A user also has the ability to save residuals and predicted values to their dataset for further analysis. The T test module allows the user to perform either a one variable or two variable Student's T test. The user can select configuration options, including the direction of the test and the confidence level of the test. The results, including the t statistic, degrees of freedom, p-value, and a confidence interval, will be displayed in a text box in the results panel.

Each module is a self contained set of R code that is dynamically added to the application at run time. What this means is that `intRo` can be easily extended by the addition of modules within the framework underlying the application. This extensibility allows for `intRo` to be customized for an individual classroom setting, tailored to the needs of a particular course.

As an example of possible customization, an instructor may wish to introduce hypothesis testing from a nonparametric framework rather than distributional. In this case, a wilcoxon rank sum test could be added easily as a module to be used instead of the T test. This addition would be fairly simple to complete. The user would fork `intRo` from http://github.com/gammarama/intRo and begin by adding six types of files to a folder within a module folder: `helper.R`, `libraries.R`, `observe.R`, `output.R`, `reactive.R`, and `ui.R`. Each of these files is written to comply with the reactive nature available in a Shiny application. Further details on the creation of new modules can be found in section \ref{modules}.

@@

The process by which `intRo` is dynamically created upon running the application through module sourcing is detailed in figure \ref{fig:app_creation_modules}. The key driver to populating \texttt{server.R} and \texttt{ui.R} is the navigation file \texttt{modules.txt}, which defines the placement of each module. The interface is then created with the following statement.

@@

# Using `intRo`

Although users can access intRo from \url{http://www.intro-stats.com}, course instructors may wish to customize their own instance.

## `intRo` package

`intRo` can be downloaded, ran, and deployed on ShinyApps.io through the use of the R package `intRo`. Currently, the package is only available on GitHub, and can be installed using the devtools package as follows:

If the intent is to use a specific instance of `intRo` where many users will access it at the same time, such as in an introductory statistics class, it may be preferable to deploy a custom instance of `intRo` to a publicly accessible URL. The package provides a function `deploy_intRo` which is a wrapper for the `deployApp` function contained in the shinyapps package. Once the shinyapps package is installed and configured, `deploy_intRo` will upload `intRo` as an application on the user's account. The function takes the same arguments as `run_intRo`, so it can be deployed with a custom selection of modules, and a customized theme. It also takes an additional argument `google_analytics`, which allows the specification of a Google Analytics tracking ID. It also takes `...` as additional arguments to be passed into the `deployApp` routine. For example, if we wished to deploy the instance of `intRo` we ran previously, we would call it like so:

## In the classroom

`intRo`'s ease of use lends itself to a natural presentation in an introductory statistics course. Statistics software typically used in introductory courses, such as JMP and Excel, have restrictive software licensing and platform requirements that can limit the software's accessibility to students. Even if these limitations can be overcome, these software programs have non-ideal interfaces for an exploratory data analysis. JMP, for instance, hides several basic regression functions such as a line of best fit behind small red arrows. Using R itself in the classroom is an appealing alternative, but suffers from some other problems. Many students in introductory statistics courses have not been exposed to programming languages and concepts, which could require a substantial overhead in order to begin a basic statistical analysis. When results need to be submitted as homework, getting the output into a document can be a frustrating and tedious experience (this experience is made much easier with new tools like rmarkdown, but this again is a bit of a learning curve for the new statistics student).

We believe that `intRo` targets these key limitations. The interface is designed to be quick, seamless, and easy to understand. Because `intRo` is web-based, the statistical analysis can be performed from nearly any modern device. R underlies the analysis, but users need not be familiar with R or programming to perform it. The focal point of the interface is the output panel, allowing the student to remain focused on the results and limiting the need to fight with the interface itself to get work done. For those students who may have a natural inclination for programming, executable code is generated to allow a local reproduction of the analysis.

The features of `intRo` were selected in order to apply to the most commonly used lesson plans in early statistics courses. The textbook "The Art and Science of Learning from Data" by Agresti and Franklin [-@agresti2012statistics] references many of the focal points of introductory statistics, and `intRo` supports performing a majority of them. Part one, titled "Gathering and Exploring Data", covers graphical summaries, measures of spread and center, and avoiding the misuse of graphics in situations they are not appropriate. `intRo` includes modules for both graphical and numerical summaries covering these topics. Furthermore, `intRo` limits the display of variables to ones that are appropriate for a chosen plot type, helping to reinforce the appropriateness of particular graphics (Table \ref{tab:plottypes}).

\begin{table*}[ht!]
\centering
\begin{tabular}{|c|c|c|}
\hline
Plot Type&X Variable Type Allowed&Y Variable Type Allowed\\
\hline
Histogram&Numeric&N/A\\
Normal Quantile Plot&Numeric&N/A\\
Scatterplot&Numeric&Numeric\\
Line Chart&Numeric&Numeric\\
Boxplot&Categorical&Numeric\\
Bar Chart&Categorical&Numeric\\
Pareto Chart&Categorical&Numeric\\
Mosaic Plot&Categorical&Categorical\\
\hline
\end{tabular}
\caption{The plot and variable types allowed in `intRo`'s graphical module. For example, when the selected plottype is histogram, the interface displays only a single variable selection box (x), and the variables selectable in this box are only numeric. If a boxplot is selected, the x variable must be categorical and the y variable must be numeric, and the list of choices updates automatically to reflect this.}
\label{tab:plottypes}
\end{table*}

Part two of this text discusses probability and sampling distributions, which is outside the current realm of `intRo`, but part three focuses on inferential statistics. The inferential statistics include confidence intervals, hypothesis testing, and comparing two variables, all supported natively in `intRo`. Finally, part four of the text discusses the association between variables, including regression, functions that `intRo` performs with ease. Some of the more advanced topics of this section are currently unsupported by `intRo`, including multiple regression.

Statistics courses with a lab component can most benefit from the use of `intRo`. As students learn the aforementioned concepts, they can practice using the tools to perform the statistical tests and summaries. They simply upload a dataset and click the module appropriate to the section of the class. Once they choose the options desired by the instructor (variables, confidence levels, etc.), they can view the results and ultimately print the output for turning in. Early labs can walk the user through the limited number of button presses and clicks necessary, while later labs can more vaguely describe the intended output in hopes that the user can easily navigate the `intRo` interface to obtain the results. For topics currently unsupported by `intRo`, the instructor can supplement its use with other web-based modules that are publicly available or create their own modules within `intRo` as detailed in section \ref{modules}.

To fully understand `intRo`, it is important to expand upon three underlying concepts. The first is the idea of modularity in the context of a Shiny application. The second is the idea of reproducibility in the context of a graphical user interface. Finally, the third is the idea of reactive programming in the context of a web application.

@@

This script is most immediately similar to standard R code. In this case, a function `cont.table` is created which, depending on the values of the parameters, ultimately returns a contingency table. One important difference from a typical R script is that each call in the script is wrapped in a function called `interpolate`. `interpolate` both executes the given R code on the server, and also writes the code executed to the script window at the bottom of `intRo`. More details on the reasons and choices behind `interpolate` can be found in section \ref{reproducibility}.

@@

## Reproducibility
Though they can make common actions quicker and interacting with software much simplier, many Graphical User Interfaces for scientific software packages come with a major drawback. That is, the actions taken by the user are typically not reproducible. Code scripting, on the other hand, allows us to share a file containing code, and another user can execute the code to reproduce the same results. In terms of the scientific process, where results can and should be scrutinized by other parties, the utility of GUIs are greatly limited.

We have designed `intRo` modules to reproduce actions taken by the user so that they can obtain R code associated with each module interacted with. When a user chooses to store the results associated with a particular module, R code yielding the visible results is automatically added to the code window at the bottom. This framework yields several advantages:

We designed the modules in such a way that the code printed to the console is the same as the code executed by the server, which makes the process of writing an `intRo` module significantly more seamless. The workhorse function for this reproducibility framework is called `interpolate`, which is based on a function proposed by Hadley Wickham [-@interpolate]. The idea of interpolation is that by using non-standard evaluation, the call can be captured, printed, and evaluated, with all the appropriate arguments substituted into the call automatically. This means that depending on the options selected in `intRo`, the corresponding arguments to functions are appropriately set for reproduction in a standard R session.

@@

@@

# Supplementary material
All code and documents related to this manuscript are available at \url{https://github.com/gammarama/intRo}.
